\usepackage[british]{babel}
\emergencystretch=20pt
\input {data}

\def \editrans {\vfill
\begin{flushright}
编辑：戴谊凡（主编），鲍里斯·伊奥姆丁，玛丽亚·鲁宾斯坦。\\
翻译：陈润，刘闽晟。
\end{flushright}}

%%% bgn language defs
%
\def \olympiad {第一届国际理论、数理及应用语言学奥林匹克竞赛}
\def \whenwher {保加利亚，波罗维茨，2013年9月8日 — 12日}

\def \pgheader {第一届 IOL:波罗维茨 '03}
\def \pontword {分}
\def \probword {题}
\def \soluword {题解}
\def \asgtword {任务}
\def \fordword #1{翻译成#1}
%
\def \famiword {family}
\def \maleword {man}
\def \femaword {woman}
\def \klukword {boy}
\def \girlword {girl}
\def \pareword {parents}
\def \kidsword {kids}
\def \lettword {letter}
%
\def \assortiR #1{\word{#1assortir} & 再次采摘 & \word{assortir} & 采摘 \\}
\def \cureR #1{\biline {#1}{curer}{清洁}}
\def \formeR #1{\word{#1former} & 改革 & \procherk \\}
\def \former #1{\word{#1former} & 再次形成 & \word{former} & 形成 \\}
\def \futer #1{\word{#1futer} & 反驳 & \procherk \\}
\def \lancer #1{\word{#1lancer} & 再次投 & \word{lancer} & 投 \\}
\def \munEreR #1{\word{#1mun\'erer} & 赔偿 & \procherk \\}
\def \partiR #1{\word{#1partir} & 分配 & \procherk \\}
%
\def \comment {\paragraph {注:}}
%
\makeatletter
\def \@alph #1{\ifcase#1\or a\or b\or c\else\@ctrerr\fi}
\makeatother
%
%%% end language defs

\begin{document}
\makepart{个人赛}

\problem {20}
%
1916年，俄国学者雅各·林茨巴赫发明了一种世界通用的书写系统，他认为所有人不论何种母语都应该能够理解。林茨巴赫称他的新语言为「超越代数」。

以下是一些林茨巴赫的语言写的句子和中文翻译:

\medskip \linzdata
{[这]父亲和[这]兄弟 (单数) 在谈论}
{[这些]巨人们在不匆忙地工作}
{[这些]孤儿们在写一封信}
{不是我们写了你}
{[这]信不是她写的}
{[这]父亲不喜欢[这]工作}
{[这]邪恶的巨人吃了[这对]父母亲}
{她没有在匆忙}

\assignment \fordword {汉语}:

\medskip \linztest


\assignment 用「超越代数」表示:

\medskip
\begin{tabular}{rl}
\birow {我的丈夫和我 (即我和[这]丈夫) 没有谈论 (过去时) 他们。}
\birow {[这些]人们在不情愿地工作。}
\birow {[这]善良的寡妇喜欢[这]没工作的侏儒。}
\birow {你们将会被谈论。}
\end{tabular}
\medskip \\
%
请解释你的答案。
\by{(克谢尼娅·吉利亚洛娃)}

\newpage
\problem {25}
%
以下是埃及阿拉伯语%
\footnote{阿拉伯语的埃及方言有大约四千五百万人使用。由于埃及显著的经济、政治和文化影响，
更重要的是广播和电视节目的巨大数目与普及，说阿拉伯语其他方言的人普遍能够理解该方言。}
的数学等式。
%
所有的加数与和，除了最后一个，都由分子分母都不大于~$10$且分母不为~$1$的分数表示:
%
\fracdata
%
\assignment 写出数字形式的等式。
\assignment 等式\quad\fractest\quad 少了一个符号。补上该符号。
\comment
字母 \wipa{\sh} 发音为英语的 \word{sh}，\wipa x 为单词 \word{loch} 中的 \word{ch}；
\wipa C 是阿拉伯语特有的辅音。元音上的横线表示长音。

\by{(戴谊凡)}

\problem {15}
%
下列是一组巴斯克语%
\footnote{在巴斯克 (西班牙的一个自治区) 以及法国境内，有超过五十万人使用巴斯克语。巴斯克语目前未被证明与任何其他语言相关。}
%
及其乱序的汉语翻译 (一些词没有给出):

\medskip\basqdata
%
\begin{quote}
二月的第一个周四；
\CJKunderline{\qquad\qquad}的最后一个周三；
十二月一日，周三；十二月的最后\underline{\qquad\qquad}；
六月九日，周日；一月二十三日，\underline{\qquad\qquad}；
十月的最后一个周六；
十一月的第三个周五；
九月\underline{\qquad\qquad}，周一；
一月二日，周五。
\end{quote}

\assignment
将原文与译文正确对应起来并填空。

\assignment \fordword {巴斯克语}:
%
\begin{quote}
十二月的第一个周一；
十一月二十九日，周六；
一月的第二周；
二月三日，周一。
\end{quote}
%
\assignment
你认为巴斯克语中星期里的日子 \word{astelehena\-}、\word{asteazkena} 和 \word{asteartea}字面翻译可能是怎样的？
\by{(亚历山大·阿尔希波夫)}

\newpage
\problem {20}
%
以下是一些使用一简化的拉丁化方案拼写的阿迪格语
\footnote{阿迪格语是属于阿布哈兹-阿迪格 (西北高加索语系) 语族。超过三十万人使用该语言，主要在阿迪格共和国 (俄罗斯联邦)。}
句子及其汉语翻译:

\medskip
\setcounter{rowcount}0
\adygdata
{他把[这]水壶放在[这]炉子里面}
{他把什么投在[这]碟子上面}
{他把[这]钱落在[这]箱子上面}
{他把[这]大锅放在[这]桌子上面}
{他把什么落在[这]凳子下面}
{他把[这]碟子放在哪里}
{他把[这]碟子投在哪里}

\assignment 请给出六、七两句更加精确的翻译。

\assignment \fordword {汉语}:

\medskip
\adygtest

\assignment \fordword {阿迪格语}:

\medskip
\begin{tabular}{rl}
\birow{他把[这]碟子放在[这]水壶下面。}
\birow{他把什么投在[这]箱子下面?}
\birow{他把什么落在[这]大锅里面?}
\end{tabular}

\assignment \fordword {阿迪格语}，写出所有译法:

\medskip
\begin{tabular}{rl}
\birow{他把[这]桌子放在哪里?}
\end{tabular}

\comment
\adygcons\ 是阿迪格语特有的辅音, \wipa@ 和 \wipa y 是阿迪格语的元音。
\by{(Yakov Testelets)}

\newpage

\problem {20}
%
下列的表格包括了法语含前缀动词和对应的无前缀动词，以及所有的中文解释。
阴影区域表示含前缀动词没有对应的无前缀动词。一些动词的前缀未给出。

\medskip
\begin{tabular}{llll}
\word{r\'eagir} & 反应 & \procherk \\
\assortiR {\underline{\quad}}
\word{recommencer} & 重新开始 & \word{commencer} & 开始 \\
\word{recomposer} & 重新编写 & \word{composer} & 编写 \\
\biline {r\'e}{concilier}{和解}
\biline {r\'e}{conforter}{安慰}
\word{recr\'eer} & 再创造 & \word{cr\'eer} & 创造 \\
\word{r\'ecr\'eer} & 娱乐 & \procherk \\
\cureR {\underline{\quad}}
\word{redire} & 再说一遍 & \word{dire} & 说 \\
\word{r\'eduire} & 减少 & \procherk \\
\word{r\'e\'editer} & 再版 & \word{\'editer} & 出版 \\
\word{refaire} & 重做 & \word{faire} & 做 \\
\formeR {\underline{\quad}}
\former {\underline{\quad}}
\futer {\underline{\quad}}
\word{r\'eincarner} & 转生 & \word{incarner} & 具体化 \\
\word{rejouer} & 继续玩 & \word{jouer} & 玩 \\
\lancer {\underline{\quad}}
\munEreR {\underline{\quad}}
\word{r\'enover} & 翻新 & \procherk \\
\word{r\'eop\'erer} & 再次操作 & \word{op\'erer} & 操作 \\
\word{repartir} & 再次离开 & \word{partir} & 离开 \\
\partiR {\underline{\quad}}
\word{r\'ep\'eter} & 重复 & \procherk \\
\biline {r\'e}{sonner}{发声}
\word{r\'ev\'eler} & 揭露 & \procherk \\
\end{tabular}
\medskip \\
%
\paragraph{任务:}
利用以上表格的信息填空。解释你的答案。
\by{(鲍里斯·伊奥姆丁)}

\editrans
\makepart{Team Contest}

\problem {35}
%
In the first millennium CE there were in Chinese Turkestan
two closely related languages, Tocharian A and Tocharian B,
which had descended from a common ancestor, Proto-Tocharian.
Here are some Proto-Tocharian words
as they have been reconstructed by scholars:

\medskip
\begin{tabular}{|ll||ll||ll|}
\wipa{\A k{\schwa}natsa} & `unreasonable' & \wipa{p{\schwa}ratsako} & `chest (breast)' & \wipa{st\A n5k{\schwa}} & `palace' \\
\wipa{\A sare} & `dry' & \wipa{r{\schwa}s{\schwa}k{\schwa}re} & `sharp' & \wipa{ts{\schwa}n5k{\schwa}r} & `top' \\
\wipa{\A st{\schwa}re} & `pure' & \wipa{sam{\schwa}} & `same' & \wipa{w{\schwa}lo} & `king' \\
\wipa{k{\schwa}r\A m{\schwa}rtse} & `black' & \wipa{s\A k{\schwa}re} & `happy' & \wipa{y{\schwa}s\A r} & `blood' \\
\end{tabular}
\medskip \\
%
And here are Tocharian A and Tocharian B words
which are descendants of the Proto-Tocharian words listed above (in no particular order):

\tochtest
%
\assignment
Determine which word belongs to which language, knowing that:
%
\begin{itemize}
\item in one of the languages some words have two variants;
\item the first word is Tocharian A.
\end{itemize}

\assignment
Allocate the following words to languages and reconstruct the Proto-Tocharian form of each pair:
%
\tochmore {tree}{red}{\lettword}

\assignment
It is thought that Tocharian B had stress
(as in English more or less).
Upon what might this hypothesis be based?

\comment
\wipa{\A} is a prolonged \wipa a, \wipa{\d s} sounds as \word{sh},
\wipa{n5} as \word{ng};
the sequence \wipa{ts} is pronounced as a single consonant,
\wipa{\schwa} is a specific Tocharian vowel.
\by{(Svetlana Burlak)}

\newpage
\problem {30}
%
When describing how personal and reflexive pronouns work in various languages,
linguists make use of the so-called subscripts---Roman letters
(typically $i$, $j$, $k$, \dots)\ which mark pronouns and some other words
in sentences. The character $*$ (asterisk) is also used.
Here are some English examples:

\begin{enumerate}
\item John$_i$ saw himself$_i$ in the mirror.
\item John$_i$ says that he$_{i/j/{}^*k}$ doesn't know Peter$_k$.
\item The boy$_i$ is playing with his$_{i/j}$ gun.
\item His$_i$ teacher$_j$'s influence in easily seen in his$_{i/{}^*j/k}$ work.
\item The girl$_i$ saw her$_{{}^*i/j}$.
\end{enumerate}

\assignment Explain the meaning of the subscripts and the asterisk.

\assignment Add subscripts (and asterisks where appropriate)
in the following sentences:

\begin{items}
\item She doesn't like this trait in herself.
\item The father took his son to his room.
\item John knows that Peter has given his book to his son.
\end{items}
\by{(Maria Rubinstein)}

\problem {35}
%
Consider the following pairs of verbs with closely related meanings:

\medskip
\begin{tabular}{ll}
\word{accuse} & \word{rebuke} \\
\word{denounce} & \word{reprehend} \\
\word{command} & \word{instruct} \\
\word{advise} & \word{guide} \\
\word{assure} & \word{convince} \\
\end{tabular}
\medskip \\
%
It is known that all verbs in the left-hand column have a certain ability
that the verbs in the right-hand column lack.

\assignment Identify the ability in question.

\assignment Find the verbs that also have this ability among the following:
\word{extort}, \word{threaten}, \word{forbid}, \word{swear}, \word{shout},
\word{approve}, \word{refuse}, \word{rob}, \word{dedicate},
\word{lose}, \word{scold}, \word{give up}, \word{demand}.

\assignment Try to find two more verbs with the same ability.
\by{(Boris Iomdin)}

\editrans
\makepart{Solutions to the Problems of the Individual Contest}

\solution

\begin{enumerate}
\item Nouns:
\begin{itemize}
\item \linzglos {work}
\begin{itemize}
\item Combinations: \linzglot {husband + wife}{brother + sister}
\item Family members are singled out by division and cancellation:
\linzglou {father}{brother}
\item Missing (deceased) family members are preceded by a minus sign:
\orphform\ = orphans' (apparently orphaned children of one and the same family).
\end{itemize}
\item $\It$ `person', $(>\It)$ `giant'.
\end{itemize}
\item Pronouns are composed of the character $\It$ or $\She$ (for feminine gender)
and the subscripts $1$ to $3$, which indicate the person.
\item The plural of nouns and pronouns is expressed by the coefficient $n$.
The plus sign plays the part of the conjunction `and'.
\item Verbs: $\talk$ `talk', $\work$ `work', $t$ `hurry',
$\scribe$ `write', $\heart$ `like, love', $\eat$ `eat'.
If what the verb denotes is absent or uncharacteristic, a minus sign expresses that:
$-\heart$ `not inclined to affection = wicked'.
(We can assume that a characteristic property is expressed by a plus sign,
hence $+\heart$ `good', a concept we need.)
\item Sentence structure:
\begin{itemize}
\item the subject is the base of the power;
\item the predicate is the exponent,
whereby negation is expressed by a minus sign ($-\heart$ `not like')
and passive voice by a radical sign ($\sqrt{\scribe}$ `be written');
additional activities can be added or subtracted
($\It_3^{\work-t}$ `he is working and doesn't hurry = he is working without haste');
\item past tense is marked by $-t$ ($\It_3^\work-t$ `he worked'),
future tense by $+t$;
\item the direct object, if there is one, follows an equals sign.
\end{itemize}
\end{enumerate}

\setcounter{rowcount}8
\assignment
%
\begin{tabular}[t]{rl}
\birow{He loves with an unrequited love (\emph{i.~e.}\ loves without being loved).}
\birow{The taciturn (\emph{or} mute) daughter will write about the father and the mother.}
\birow{You (sg.\ fem.)\ worked quickly (\emph{or} hastily) and silently.}
\birow{The letter was eaten by the hungry sister.}
\end{tabular}

\assignment \linzdone

\solution
%
All Arabic words in the problem are made according to one of the patterns
\wipa{1a2a3t}, \wipa{i12\A3}, \wipa{1u23} and \wipa{1u23\E n}
(whereby words using the first and the second pattern always come together in this order
and words using the other two patterns occur on their own).
In these patterns \wipa{1-2-3} is one of the triples of consonants
\wipa{r-b-C}, \wipa{s-b-C}, \wipa{s-d-s}, \wipa{t-l-t},
\wipa{t-m-n}, \wipa{t-s-C}, \wipa{x-m-s}, \wipa{C-\sh-r}.
Let us assume that the consonant triples correspond to numbers between $1$ and $10$
and the arrangements of the vowels indicate certain functions, in particular,
\wipa{1a2a3t} \wipa{i1$'$2$'$\A3$'$} is either $\frac n{n'}$ or $\frac {n'}n$
(and in either case $\egar{xamast ixm\A s}=\frac nn=1$),
and $\egar{1u23} = \frac in$ and $\egar{1u23\E n} = \frac jn$,
for some as yet unknown $i$ and $j$.

From equality (5) we see that \wipa{s-b-C} and \wipa{x-m-s} are $5$ and $7$
(in one order or the other), and from $\frac j5+\frac j7=\frac{(7+5)j}{35}=\frac{24}{35}$
it follows that $j=2$, that is, $\egar{1u23\E n} = \frac 2n$.
Since \wipa{1u23} is shorter than \wipa{1u23\E n}, we can assume
that this pattern corresponds to a more basic function, and the only candidate for such a one is $\frac 1n$.

From (1) it follows that \wipa{t-l-t} is $3$
(and that the numerator precedes the denominator in the Arabic fractions).
From (4) we see that \wipa{t-m-n} is greater than \wipa{s-b-C} by one.
From (3) it follows that $3\egar{s-d-s} = 2\egar{t-s-C}$.
Thus \wipa{t-s-C} is divisible by three. Since the value $3$ is already taken,
\wipa{t-s-C} and \egar{s-d-s} are either $6$ and $4$ or $9$ and~$6$, respectively,
and \wipa{t-m-n}, \wipa{s-b-C} and \wipa{x-m-s} are respectively $8$, $7$ and $5$.

We have yet to use equality (2).
Letting \wipa{s-d-s} be equal to $4$ gets us nowhere
($\frac73 + \frac14 = \frac{31}{12}$ can't be reduced to a fraction
with a numerator and denominator between $1$ and $10$),
consequently $\egar{s-d-s}=6$,
and $\frac73 + \frac16 = \frac{15}6 = \frac52 = \frac{10}4 = \egar{C-\sh-r}/\egar{r-b-C}$.
(The root \wipa{r-b-C} `$4$' is the source of the word \word{ruba'i} `quatrain',
used also in English.)

\assignment
%
\fracdone

\assignment
%
$\egar{rubC} + \egar{Ca{\sh}art its\A C} = \frac14 + \frac{10}9 = \frac{49}{36}$
and $\egar{sabaCt isd\A s} = \frac76$.
Thus either $\sqrt{\egar{rubC} + \egar{Ca{\sh}art its\A C}} = \egar{sabaCt isd\A s}$
or, perhaps, $\egar{rubC} + \egar{Ca{\sh}art its\A C} = (\egar{sabaCt isd\A s})^2$
(if we don't consider brackets to be a sign).

\newpage\solution
%
There are two types of English expressions in the problem:
some (I) consist of a date, a month and a day of the week,
others (II) name the number of the day of the week within the month instead of the date.
The word order in the Basque expressions of type (I) is \<month> \<date>, \<day of the week>,
whilst in type (II) it is \<month> \<number of the day> \<day of the week>.
The last word ends in \word{-a},
whereas the preceding words have no final \word{-a}
(except for the word \word{hogeita}, which means `20' in compound numerals).
The element \word{-garren} forms ordinal numbers.
The word \word{astea} is not a name of a day of the week
(six of those we have seen in examples 1--10, the seventh occurs in Assignment~3).
Since Assignment 2 features the word `week', we can guess
that this is the meaning of the word \word{astea}.

\assignment
%
\begin{tabular}[t]{ll}
\word{urtarrilaren hogeita hirugarrena, larunbata} & \word{the 23rd of January, \underline{Saturday}} \\
\word{abenduaren azken astea} & \word{the last \underline{week} of December} \\
\word{otsailaren lehenengo osteguna} & \word{the first Thursday of February} \\
\word{ekainaren bederatzigarrena, igandea} & \word{the ninth of June, Sunday} \\
\word{abenduaren lehena, \underline{asteazkena}} & \word{the first of December, Wednesday} \\
\word{irailaren azken asteazkena} & \word{the last Wednesday of \underline{September}} \\
\word{azaroaren hirugarren ostirala} & \word{the third Friday of November} \\
\word{urriaren azken larunbata} & \word{the last Saturday of October} \\
\word{irailaren lehena, astelehena} & \word{\underline{the first of} September, Monday} \\
\word{\underline{urtarrilaren} bigarrena, ostirala} & \word{the second of January, Friday} \\
\end{tabular}

\assignment
%
\basqmore
{the first Monday of December}{the 29th of November, Saturday}
{the second week of January}{the third of February, Monday}

\assignment
%
\word{Astelehena} `Monday', \word{asteazkena} `Wednesday';
\word{asteartea}, the only day of the week not found in in Assignment~1, is `Tuesday'.
All three names are formed from the word \word{aste} `week'.
\word{Astelehena} means literally `first (day) of the week',
\word{asteazkena} `last (day) of the week'.
Tuesday's Basque name can be translated more or less
as `day in the middle of the week'.

No one knows for sure why Basque calls Wednesday `last day of the week'.
In Basque dialects other variants of the names of the days of the week are also found,
including loans from Romance languages.

\newpage
\solution
%
The Adyghe sentences have the following structure:

\medskip
\adyganal {He $V$ & $X$ & $P$ $Y$}{What does he $V$ && $P$ $Y$}{Where does he $V$ & $X$}
\medskip \\
%
where $X$ and $Y$ are nouns, $V$ is a verb (or its stem)
and $P$ is, in English, one of the prepositions \word{into}, \word{onto} or \word{under}
and in Adyghe it is one of the prefixes \wipa{d-}, \wipa{tyr-} or \wipa{\d{\cj}-}.
As the third schema shows, the Adyghe locative prefix
may not correspond to anything in the natural (but imprecise) English translation.

\assignment
%
We specify (at the expense of naturalness):

\setcounter{rowcount}5
\medskip
\begin{tabular}{rl}
\birow{Under what does he put the plate?}
\birow{Onto what does he throw the plate?}
\end{tabular}

\assignment
%
\begin{tabular}[t]{rl}
\birow{He throws the stool into the stove.}
\birow{Where (into what) does he drop the money?}
\end{tabular}

\assignment \adygdone

\assignment
%
\adygmore {Into what does he put the table}{Onto what does he put the table}{Under what does he put the table}

\vfill
\solution

\frendone \medskip \\
%
The table features verbs with two different prefixes: \word{re-} and \word{r\'e-}.
All verbs with \word{re-} indicate a repetition
or a renewal of the action named by the verb without a prefix.
Contrariwise, if the prefix is \word{r\'e-},
then the corresponding prefixless verb either doesn't exist
or means the same thing as the prefixed one does.
The verbs whose stems begin with vowels are an exception:
the prefix they take is \word{r\'e-}
regardless of the existence and the meaning of a corresponding prefixless verb.
There are other exceptions from this rule in French,
but on the whole it is fairly reliable.

\comment
The vowel in the prefix \word{r\'e-} is not unlike the first vowel in \word{raider},
whereas the one in the prefix \word{re-} bears a certain similarity to the second,
and needs to be fortified when it finds itself next to another vowel.

\editrans
\makepart{Solutions to the Problems of the Team Contest}

\solution
%
\assignment

\medskip
\tochdone
\medskip \\
%
The first pair gives the correspondence \wipa{\d st}---\wipa{st}.
This determines unambiguously the second pair (or triple, rather), whence we learn
that Tocharian B has kept the final vowels (except for the `specific' one) and Tocharian A has lost them.
Consequently all words  with retained final vowels are Tocharian B
and their counterparts with lost final vowels are Tocharian A.
This allows the following conclusions to be made:
In Tocharian A the `specific' vowel falls out before a vowel that is retained
and is retained before one that is lost;
\wipa a, long or short, is preserved without change.
In Tocharian B the `specific' vowel can become \wipa a, \wipa{\schwa} or nothing
and both {\wipa a\/}s can become either \wipa a or \wipa{\A}.
This determines the remaining pairs.

\assignment
%
(a) A \wipa{\d st{\A}m}, B \wipa{st{\A}m} `tree' $<$ *\wipa{st{\A}m{\schwa}};
(b) A \wipa{rt{\schwa}r}, B \wipa{ratre} `red' $<$ *\wipa{r{\schwa}t{\schwa}re};
(c) A \wipa{p{\schwa}rs}, B \wipa{parso} `\lettword' $<$ *\wipa{p{\schwa}rso}.
%
In the reconstruction the `specific' vowel is not inserted in clusters of the type
`sonant + obstruent' and the cluster \wipa{st},
nor is it added after final \wipa r.

\assignment
It is assumed that under stress *\wipa{\schwa} $>$ \wipa a, *\wipa{a/\A} $>$ long \wipa{\A},
whereas without stress *\wipa{\schwa} $>$ nothing or \wipa{\schwa} (as in Tocharian A),
*\wipa{a/\A} $>$ short \wipa a.

\solution
%
\assignment The subscripts mark the participants in the situation (the persons mentioned in the sentence).
Identical letters mean identical individuals, different letters mean different individuals.
In this way it is shown which pronoun can refer to which noun.
If a pronoun can refer to more than one noun,
all possible subscripts are given, separated by slashes.
If a pronoun can refer to an individual not mentioned in the sentence,
a letter is used that doesn't mark any other word in the same sentence
(e.~g., \word{he} in (2) may be someone other than John or Peter, let's say Bill, if he exists at all).
An asterisk next to a letter indicates
that the pronoun can't refer to the noun with this subscript.

\assignment
\begin{items}
\item She$_i$ doesn't like this trait in herself$_i$.
\item The father$_i$ took his$_{i/{}^*j/k}$ son$_j$ to his$_{i/j/k/l}$ room.
\item John$_i$ knows that Peter$_j$ has given his$_{i/j/l}$ book to his$_{i/j/k/l/m}$ son$_k$.
\end{items}

%\newpage
\solution
%
\assignment
%
The left column contains what are technically known as performative verbs.
(The concept of performativity was introduced in 1965 by the English philosopher John Austin.) 
They are different from other verbs in that
the action they name can be performed by their use, rather than simply described.
So the words \word{`I accuse you of murder'}
all by themselves constitute an accusation;
the words \word{`I denounce you as an impostor'}, a denunciation;
\word{`I command you to report to the headquarters at once'}, a command;
\word{`I advise you not to go there'}, advice;
\word{`I assure you that this problem is not so hard'}, assurance.
Performativity is a rather peculiar property;
as the statement of the problem shows,
even verbs with very similar meanings can differ in its presence or absence
(one can't very well say \word{`I hereby reprehend your cowardice'}
of \word{`I convince you that this is the correct solution'}).

\assignment
%
These are the verbs \word{forbid} (\word{`I forbid leaving the room before the class is over'}),
\word{swear} (\word{`I swear to cheat no more'}),
\word{approve} (\word{`I approve of your decision'}),
\word{refuse} (\word{`I refuse to try to solve this problem'}),
\word{dedicate} (\word{`I dedicate this book to my parents'}),
\word{give up} (\word{`I can't do this problem, I give up'}),
\word{demand} (\word{`I demand to be told how this problem is to be solved'}).

\assignment For example,
\word{thank} (\word{`I thank you for the clarification'}),
\word{congratulate} (\word{`I congratulate you on your success'}).

\editrans
\end{document}
